Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2023-05-11 analysis/paper/paper.qmd
stash

The introduction of new stone artifact technologies marked a major transition in the Korean Paleolithic, from the Early (approx.350~40 ka) to the Late Paleolithic periods (approx.40~12 ka). The transition includes blade technology, stemmed points, end scrapers, burins, denticulates, etc. [@bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean; @seong2016eastern]. Stemmed points are considered to be the first evidence of a suite of new technologies defining the Late Paleolithic period in this region [@seong2016eastern; @seong2008tanged]. This is related to the fact that stemmed points appear to originate from Korea and spread throughout Northeast Asia, and they have a close association with mobility, site formation, and occupation diversity [@o2018origins; @Chong_2021; @park2022did]. Despite the importance of stemmed points, only a few studies to date have examined their likely uses. Previous work mostly discussed their origin, the chronology of the Korean Late Paleolithic, and their relationship with the Japanese archipelago [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].

The purpose of this study is to examine the possible uses of stemmed points to understand what role they may have played in the technological transition from the Early to the Late Paleolithic in Korea. We use the tip cross-sectional area (TCSA) metric to infer weapon-use strategies based on comparison with other archaeological and ethnographic assemblages, assuming that different ranges of TCSA values correspond to different weapon types [@lombard2021variation]. We then explore the relationship between the TCSA range and raw materials, artifact size and discard location, and how these changed over time in Korea. Our main questions are: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point uses? We examine possible links between the roles of stemmed points and environmental change, especially the Last Glacial Maximum, or population dynamics during the Late Paleolithic period. To understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types as part of their adaptation to their socio-environmental circumstances.

@@

Stemmed points (Sumbejjirugae in Korean) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two ridges that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made). Slight retouch is typically performed on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft. Elsewhere in the world these types of artifacts are often called "tanged points", but we prefer "stemmed points" to distinguish them from Bronze Age stone projectile points known as "tanged points" in Korea [@park2022did]. Understanding the appearance of stemmed points is relevant to general questions about the direction of projectile technology, the technological transition into the Late Paleolithic, and relationships between Korea and adjacent regions in East Asia, such as Japan.

@@

Since the first appearance of stemmed points defines the beginning of the Korean Late Paleolithic, investigating their origins is critical to understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern]. The debate around the origin of stemmed points can be summarized into two competing models [@bae2013early;@lee2016patterns] : (1) a 'heterogenic' migration [@bae2010origin], and (2) an *in situ* evolution model [@Seong_2006]. The migration model claims that the new blade industry - including stemmed points - and the earlier coarse flake tradition - including large cores, polyhedrons, choppers, and even handaxes - came from different origins. These are (1) a Northern route: Siberia, Mongolia, or other regions of northeast China, and (2) a Southern route: southern China) as the result of a continuing influx of modern human migration from two routes. The *in situ* model argues that stemmed points and other Late Paleolithic technologies, such as blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence. The difference between the two models comes from how they explain a few early sites and stemmed points made out of flakes. The *in situ* model argues that stemmed points appeared before blades, and identifies early sites such as Bonggok, Songamri, Yonghodong, and Hwadaeri as evidence that the Korean Late Paleolithic began with the emergence of stemmed points. In contrast, the migration model contends that the Late Paleolithic began with the introduction and widespread use of blade technology, similar to the traditional definition of the Late or Upper Paleolithic in Europe and the Later Stone Age in Africa. [@seong2016eastern]. Besides these two main models, there are combination of migration and trade interaction model [@bae2012nature] and complex and non-directional model [@lee2016patterns].

@@

The tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabilistically discriminate between likely weapon-tip types, such as spear-thrower (a.k.a. atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins; @lombard2022standardized; @metz2023bow; @lombard2023neanderthal; @lombard2022tip]. It is critical to note that the TCSA metric alone cannot unambiguously determine artifact function; it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips. The TCSA metric represents the part of the tool that cuts the target's hide and is related to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation]. This method was first proposed by @hughes1998getting and validated by @shea2006origins through the comparison of archaeological examples with ethnographically collected samples of known use. @shea2009impact also applied the approach to compare projectile points from Africa, the Levant, and Europe, claiming that projectile weapons first appeared in Africa.

@@

After the first discovery at the Seokjangri site in the 1960s, more than 450 stemmed points have been found at over 30 sites throughout South Korea [@Sohn_1967; @lee2019were; @Chong_2021]. Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017]. Among these stemmed points, we selected those that retained their widest and thickest parts. We included stemmed points discovered during field surveys as well as those found at sites that were never dated but were associated with other Late Paleolithic artifacts. Applying these sampling criteria resulted in a sample of `r total_sp` stemmed points from `r total_ass` assemblages unearthed from `r total_site` sites spanning the period 44-10 ka (@fig-map). The dimensions of the `r total_sp` stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of local museums and archaeological institutions in Korea.

@@

To answer our research questions about the likely uses for stemmed points, we calculated TCSA values for the stemmed points in our sample. We also explored the interaction of TCSA values with raw materials and artifact size, using weight as a proxy. Because the shape of stone artifacts is highly influenced by the raw materials [@mcpherron2014experimental], we assumed that raw materials may be highly correlated with tool size. We investigated radiocarbon ages associated with the points in order to determine temporal patterns in TCSA values. We separated the research time period into three phases, based on the major climate event during the Late Pleistocene, the Last Glacial Maximum (LGM). Previous research on the LGM climate in Korea, using age-controlled pollen records [@yi2010vegetation] and computational simulation models [@kim2015last;@park2022did], indicates that the climate was colder and drier than the preceding period. Surface temperature cooling ranged from 5 to 6°C, and there was a precipitation decrease of approximately 14%. Using this chronology, we examined the distribution of TCSA values before, during, and after the LGM. We then explored the relationship between TCSA values and the location of assemblages by comparing the distribution of TCSA values across ecological and vegetation zones.

```{r}
#| label: tbl-tcsa-ranges
#| tbl-cap: "TCSA ranges from Lombard et al. (2022)"

@@

read_csv(here::here("analysis/data/raw_data/tcsa-range-from-lombard-2022.csv")) %>% 
  knitr::kable(caption = "TCSA ranges from Lombard et al. (2022)")
```

@@

The vegetation zones are based on Yi and Kim’s [-@yi2010vegetation] classification of South Korea into three zones: Central Temperate Zone (CT), South Temperate Zone (ST), and Subtropical-warm Temperate Zone (SWT). These divisions are based on Yim and Kira's [-@yim1975distribution] forest vegetation map, defined by recent temperature and precipitation values.

@@

```{r}
#| label: fig-tcsa-all-sp
#| fig-cap: "Distribution of TCSA values for all Korean stemmed points in the current dataset. The shaded boxes in color indicate TCSA ranges for different weapon types based on @tbl-tcsa-ranges."
#| fig-width:   3
#| fig-height:  2

@@

ggplot(tcsa,
        aes(0.3, 
            min_tcsa)) +
  # show regions for tools from Lomabrd et al. 2022
  # https://www-sciencedirect-com.offcampus.lib.washington.edu/science/article/pii/S0277379121002286
  # https://assets.researchsquare.com/files/rs-2016964/v1/00d381a3-3ecf-4d5b-b2ec-89a41227bbd5.pdf?c=1662475267
    annotate("rect", 
           xmin = 0.1,
           xmax = 0.2,
           ymin = 4,
           ymax = 18,
           fill = "green",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Poisoned arrow tips",
               hjust = 0,
               lineheight = 0.75,
           x = point_function_group_label_y,
           y = (18-4) /2 + 4) +
      annotate("rect", 
           xmin = 0.2,
           xmax = 0.3,
           ymin = 17,
           ymax = 47,
           fill = "blue",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Un-poisoned arrow tips",
               hjust = 0,
               lineheight = 0.75,
           x = point_function_group_label_y,
           y = (47-17) /2 + 17) +
      annotate("rect", 
           xmin = 0.3,
           xmax = 0.4,
           ymin = 40,
           ymax = 76,
           fill = "orange",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Dart tips",
               hjust = 0,
           x = point_function_group_label_y,
           y = (76-40) /2 + 40 - 5) +
      annotate("rect", 
           xmin = 0.4,
           xmax = 0.5,
           ymin = 42,
           ymax = 90,
           fill = "purple",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Javelin tips",
               hjust = 0,
           x = point_function_group_label_y,
           y = (90-42) /2 + 42 ) +
     annotate("rect", 
           xmin = 0.5,
           xmax = 0.6,
           ymin = 80,
           ymax = 200,
           fill = "yellow",
           alpha = 0.2) +
      annotate("text", 
               label = "Stabbing spear tips",
               size = point_function_group_label_size,
               hjust = 0,
               lineheight = 0.75,
           x = point_function_group_label_y,
           y = (200-80) /2 + 80 ) +
  geom_boxplot(alpha = 0.1) +
  geom_quasirandom(alpha = 0.4,
                   size = 0.7) +
  coord_cartesian(xlim = c(-0.1,1)) +
  theme_bw(base_size = 6) +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()
        ) +
  labs(y = "TCSA")

@@

############ explore tcsa with MODERN eco_regions #######
library(sf)
veg_all <- read_sf(here("analysis/data/gis_data/eco_re_two.shp"))
veg_all <- st_transform(veg_all, crs = st_crs(4326))
# ggplot() +
#   geom_sf(data = veg_all) +
#   coord_sf()
pts_all <- st_as_sf(korean_archaeological_site_locations_tcsa,
                coords = c("long_dd.y", "lat_dd.y"),
                crs = st_crs(veg_all))
# ggplot() +
#   geom_sf(data = pts_all)
sf_use_s2(FALSE) # to prevent error. Check here for more info: https://github.com/r-spatial/sf/issues/1762
points_in_veg_zones_all <- 
st_join(veg_all, 
        pts_all, 
        join = st_contains)

eco_plot <- ggplot() +
  geom_sf(data = veg_all, aes(fill = "grey")) +
   geom_sf_text(data = veg_all, aes(label = ecoregion), size =1.2)+
  geom_sf(data = pts_all, aes(color = min_tcsa), size = 0.8) +
  theme_minimal() +
  scale_fill_brewer(palette="Greys") +
  labs(color="TCSA") +
  guides(fill = "none") +
  theme(axis.text.x = element_text(angle = 45, size = 4.5))+
  theme(axis.text.y = element_text(size = 4.5))+
  theme(axis.title.x = element_blank(),
          axis.title.y = element_blank()) +
  theme(legend.position = "none")

@@

# ggplot() +
#   geom_sf(data = veg_forest) +
#   coord_sf()
pts_forest <- 
  st_as_sf(korean_archaeological_site_locations_tcsa,
                coords = c("long_dd.y", "lat_dd.y"),
                crs = st_crs(veg_forest))
# ggplot() +
#   geom_sf(data = pts_forest)
#sf_use_s2(FALSE)
points_in_veg_zones_forest <- 
st_join(veg_forest, 
        pts_forest, 
        join = st_contains)
for_plot <- ggplot() +
  geom_sf(data = veg_forest, aes(fill = "grey")) +
  geom_sf_text(data = veg_forest, aes(label = Zone), size =1.2)+
  geom_sf(data = pts_forest, aes(color = min_tcsa), size = 0.8) +
  theme_minimal()+
  scale_fill_brewer(palette="Greys") +
  labs(color="TCSA") +
  guides(fill = "none") +
  theme(axis.text.x = element_text(angle = 45, size = 4.5)) +
  theme(axis.text.y = element_text(size = 4.5)) +
  theme(axis.title.x = element_blank(),
          axis.title.y = element_blank()) +
  theme(legend.title = element_text(size= 6),
        legend.key.height = unit(4.5, 'mm'),
        legend.key.width = unit(2, 'mm'),
        legend.text = element_text(size = 4))

@@

To explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones. Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of the vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of the eco-region map. We observed that there are statistically significant differences in the distribution of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`). The Tukey's HSD results show that TCSA values from the South Temperature Zone (ST) are significantly different from the other two vegetation zones (@fig-gis-tcsa-sites panel C). Looking into this further, we see that there are only six artifacts from the ST zone, with only three of these having TCSA values above 175. The other three have TCSA values lower than 100, similar to the other zones (@fig-gis-tcsa-sites panel B). Given the small number of artifacts with extreme TCSA values in the ST zone, we hesitate to conclude that this is an archaeologically significant pattern of different TCSA values. The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values; this was likely due to the small number of points in several zones (@fig-gis-tcsa-sites panel F). Overall, our results indicate no clear patterns in the distribution of TCSA values across vegetation and eco-regional zones.

@@

```
```{r}

@@

# Use reduce() to iteratively join the data frames
tcsa_all <- reduce(dfs, left_join, by = join_col) %>%
            dplyr::select(age_ka.x, min_tcsa, Assemblage.x, raw_materials_top.x, ecoregion, Zone) %>%             mutate(mis_stage = case_when (age_ka.x > 26.5 ~ 'Before' , age_ka.x > 19 ~ 'LGM', age_ka.x < 19 ~ 'After'))

@@

plot_grid(  tcsa_all_stage_plot,
  tcsa_all_raw_plot,
tcsa_all_ass_plot,
 tcsa_all_veg_plot, 
  tcsa_all_geo_plot,
ncol =2,
label_size = 1)

@@

By comparing the results with the TCSA ranges from other archaeological and ethnographic cases (@tbl-tcsa-ranges), we were able to examine what role stemmed points played in the technological transition during the Korean Late Paleolithic. Our main questions were: What were the best-fit ballistic probabilities for stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point uses?

@@

Our results show that TCSA values vary between assemblages with few discernible temporal patterns in function (@fig-tcsa-radio-carbon panel A). Contrary to our prediction, the earliest stemmed points do not have more variable TSCA values than subsequent ones. The most striking temporal pattern is simply that stemmed points were produced and discarded most frequently before the LGM (@fig-tcsa-radio-carbon panel B). While we predicted the greatest variation in TCSA values during the LGM, our results show the opposite: variability decreased during the LGM (CV = `r cv_lgm`), in comparison to the previous period (CV = `r cv_bef`). This could be explained by the overall decline in stemmed point usage, perhaps due to a retreat into refugia during the LGM, rather than increased mobility into unfamiliar areas.

@@

Considering the importance of stemmed points for the technological transition during the Late Paleolithic in Korea, our research examines the likely uses for stemmed points by asking three research questions: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point use? We applied the tip cross-sectional area (TCSA) metric because it has been used as a ballistically relevant standard to discriminate different likely use classes of projectile points and it requires only a few measurements on a stone artifact (i.e., maximum width and thickness) to compute the metric. We calculated TCSA for a total of `r min_n_points` stemmed points from `r n_ass` assemblages excavated from `r n_sites` sites. Then we examined the TCSA values with other variables, including raw materials, weight, radiocarbon dates, and site locations. Drawing on evolutionary theory, we premised that the way of using stemmed points likely reflects the socio-environmental circumstances that people encountered and managed. We examined the possible impact of LGM on stemmed points and their distribution across eco-regional zones.

>>>>

The introduction of new stone artifact technologies marked a major transition in the Korean Paleolithic, from the Early to the Late Paleolithic periods. The transition includes blade technology, stemmed points, end scrapers, burins, denticulates, etc. [@bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean; @seong2016eastern]. Stemmed points are considered to be the first evidence of a suite of new technologies defining the Late Paleolithic period in this region [@seong2016eastern; @seong2008tanged]. This is related to the fact that stemmed points appear to originate from Korea and spread throughout Northeast Asia, and they have a close association with mobility, site formation, and occupation diversity [@o2018origins; @Chong_2021; @park2022did]. Despite the importance of stemmed points, only a few studies to date have examined their likely uses. Previous work has mostly discussed their origin, the chronology of the Korean Late Paleolithic, and their relationship with the Japanese archipelago [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].

The purpose of this study is to examine the possible uses of stemmed points in order to understand what role stemmed points may have played in the technological transition from Early to the Late Paleolithic in Korea. We use the tip cross-sectional area (TCSA) metric to infer weapon-use strategies based on comparison with other archaeological and ethnographic assemblages, assuming that different ranges of TCSA values correspond to different weapon types [@lombard2021variation]. We then explore the relationship between the TCSA range and raw materials, artifact size and discard location, and how these changed over time in Korea. Our main questions are: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point uses? We examine possible links between the roles of stemmed points and environmental change, especially the Last Glacial Maximum, or population dynamics during the Late Paleolithic period. In order to understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types as part of their adaptation to their socio-environmental circumstances.

@@

Stemmed points (Sumbejjirugae in Korean) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two arises that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made). Slight retouch is typically performed on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft. Elsewhere in the world these types of artifacts are often called "tanged points", but we prefer "stemmed points" here to distinguish them from Bronze Age stone projectile points which have long been called "tanged points" in Korea [@park2022did]. Understanding the appearance of stemmed points is relevant to general questions about the origins of projectile technology, the technological transition of the Late Paleolithic, and relationships between Korea and adjacent regions in East Asia, such as Japan.

@@

Since stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origins is critical to understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern]. The debate around the origin of stemmed points can be summarized into two competing models: (1) a 'heterogenic' migration [@bae2010origin], and (2) an *in situ* evolution model [@seong2009emergence]. The migration model claims that the new blade industry - including stemmed points - and the earlier coarse flake tradition - including large cores, polyhedrons, choppers, and even handaxes - came from different origins. These are (1) a Northern route: Siberia, Mongolia, or other regions of northeast China, and (2) a Southern route: southern China) as the result of a continuing influx of modern human migration from two routes.

The *in situ* model argues that stemmed points and other Late Paleolithic technologies, such as blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence. The difference between the two models comes from how they explain a few early sites that contain stemmed points but no blades and blade cores. While the *in situ* model claims that stemmed points appeared before blades by acknowledging those early sites, including Bonggok, Songamri, Yonghodong, and Hwadaeri, the migration model supports the traditional definition of the Late or Upper Paleolithic in Europe and Africa that started with the emergence and establishment of blade technology [@seong2016eastern]. <!-- BM: I think the last two sentences of this paragraph are unclear and confusing. I expect others will also struggle to follow the logic here and recommend your expand and clarify in detail -->

@@

Tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabilistically discriminate different likely use classes of projectile armatures, such as spear-thrower (a.k.a. atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins; @lombard2022standardized; @metz2023bow; @lombard2023neanderthal]. It is critical to note that the TCSA metric alone cannot unambiguously determine artifact function; it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips. The TCSA metric represents the part of the tool that cuts the target's hide and is related to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation]. This method was first proposed by @hughes1998getting and validated by @shea2006origins through the comparison of archaeological examples with ethnographically collected samples of known use. @shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe, claiming that projectile weapons first appeared in Africa.

@@

After the first discovery at the Seokjangri site in the 1960s, more than 450 stemmed points have been found at over 30 sites throughout South Korea [@Sohn_1967; @lee2019were; @Chong_2021]. Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017]. Among these stemmed points, we selected those that retained their widest and thickest parts. We included stemmed points discovered during field surveys as well as those found at sites that were never dated but were associated with other Late Palaeolithic artifacts. Applying these sampling criteria resulted in a sample of `r total_sp` stemmed points from `r total_ass` assemblages unearthed from `r total_site` sites spanning the period 44-10 ka (@fig-map). The dimensions of the `r total_sp` stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of local museums and archaeological institutions in Korea.

@@

To answer our research questions about the likely uses for stemmed points, we calculated TCSA values for the stemmed points in our sample. We also explored the interaction of TCSA values with raw materials and artifact size, using weight as a proxy. Because the shape of stone artifacts is highly influenced by the raw materials [@mcpherron2014experimental], we assumed that raw materials may be highly correlated with tool size. We investigated radiocarbon ages associated with the points in order to determine temporal patterns in TCSA values. We separated the research time period into three phases based on the major climate event during the Late Pleistocene, the Last Glacial Maximum (LGM), and examined the distribution of TCSA values before, during, and after the LGM. We explored the relationship between TCSA values and the location of assemblages by comparing the distribution of TCSA values across ecological and vegetation zones.

```{r}
#| label: tbl-tcsa-ranges
#| tbl-cap: "TCSA range from Lombard (2021 & 2022)"

@@

read_csv(here::here("analysis/data/raw_data/tcsa-range-from-lombard-2022.csv")) %>% 
  knitr::kable(caption = "TCSA range from Lombard (2021 & 2022)")
```

@@

The vegetation zones are based on @yi2010vegetation's classification of South Korea into three zones: Central Temperate Zone (CT), South Temperate Zone (ST), and Subtropical-warm Temperate Zone (SWT). These divisions are based on @yim1975distribution's forest vegetation map, defined by recent temperature and precipitation values.

@@

```{r}
#| label: fig-tcsa-all-sp
#| fig-cap: "Distribution of TCSA values for all Korean stemmed points in the current dataset. Shaded boxes indicate TCSA ranges for various weapon types."
#| fig-width:   3
#| fig-height:  2

@@

ggplot(tcsa,
        aes(0.3, 
            min_tcsa)) +
  # show regions for tools from Lomabrd 2021 & 2022
  # https://www-sciencedirect-com.offcampus.lib.washington.edu/science/article/pii/S0277379121002286
  # https://assets.researchsquare.com/files/rs-2016964/v1/00d381a3-3ecf-4d5b-b2ec-89a41227bbd5.pdf?c=1662475267
    annotate("rect", 
           xmin = 0.1,
           xmax = 0.2,
           ymin = 4,
           ymax = 18,
           fill = "green",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Poisoned arrow tips",
               hjust = 0,
               lineheight = 0.75,
           x = point_function_group_label_y,
           y = (18-4) /2 + 4) +
      annotate("rect", 
           xmin = 0.2,
           xmax = 0.3,
           ymin = 17,
           ymax = 47,
           fill = "blue",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Un-poisoned arrow tips",
               hjust = 0,
               lineheight = 0.75,
           x = point_function_group_label_y,
           y = (47-17) /2 + 17) +
      annotate("rect", 
           xmin = 0.3,
           xmax = 0.4,
           ymin = 40,
           ymax = 76,
           fill = "orange",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Dart tips",
               hjust = 0,
           x = point_function_group_label_y,
           y = (76-40) /2 + 40 - 5) +
      annotate("rect", 
           xmin = 0.4,
           xmax = 0.5,
           ymin = 42,
           ymax = 90,
           fill = "purple",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Javelin tips",
               hjust = 0,
           x = point_function_group_label_y,
           y = (90-42) /2 + 42 ) +
     annotate("rect", 
           xmin = 0.5,
           xmax = 0.6,
           ymin = 80,
           ymax = 200,
           fill = "yellow",
           alpha = 0.2) +
      annotate("text", 
               label = "Stabbing spear tips",
               size = point_function_group_label_size,
               hjust = 0,
               lineheight = 0.75,
           x = point_function_group_label_y,
           y = (200-80) /2 + 80 ) +
  geom_boxplot(alpha = 0.1) +
  geom_quasirandom(alpha = 0.4,
                   size = 0.7) +
  coord_cartesian(xlim = c(-0.1,1)) +
  theme_bw(base_size = 6) +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()
        ) +
  labs(y = "TCSA")

@@

############ explore tcsa with MODERN eco_regions #######
library(sf)
veg_all <- read_sf(here("analysis/data/gis_data/eco_re_two.shp"))
veg_all <- st_transform(veg_all, crs = st_crs(4326))
# ggplot() +
#   geom_sf(data = veg_all) +
#   coord_sf()
pts_all <- st_as_sf(korean_archaeological_site_locations_tcsa,
                coords = c("long_dd.y", "lat_dd.y"),
                crs = st_crs(veg_all))
# ggplot() +
#   geom_sf(data = pts_all)
sf_use_s2(FALSE) # to prevent error. Check here for more info: https://github.com/r-spatial/sf/issues/1762
points_in_veg_zones_all <- 
st_join(veg_all, 
        pts_all, 
        join = st_contains)
eco_plot <- ggplot() +
  geom_sf(data = veg_all, aes(fill = "grey")) +
   geom_sf_text(data = veg_all, aes(label = ecoregion), size =1.5)+
  geom_sf(data = pts_all, aes(color = min_tcsa), size = 1.1) +
  theme_minimal() +
  scale_fill_brewer(palette="Greys") +
  labs(color="TCSA") +
  guides(fill = "none") +
  theme(axis.text.x = element_text(angle = 45, size = 4.5))+
  theme(axis.text.y = element_text(size = 4.5))+
  theme(axis.title.x = element_blank(),
          axis.title.y = element_blank()) +
  theme(legend.position = "none")

@@

# ggplot() +
#   geom_sf(data = veg_forest) +
#   coord_sf()
pts_forest <- 
  st_as_sf(korean_archaeological_site_locations_tcsa,
                coords = c("long_dd.y", "lat_dd.y"),
                crs = st_crs(veg_forest))
# ggplot() +
#   geom_sf(data = pts_forest)
#sf_use_s2(FALSE)
points_in_veg_zones_forest <- 
st_join(veg_forest, 
        pts_forest, 
        join = st_contains)
for_plot <- ggplot() +
  geom_sf(data = veg_forest, aes(fill = "grey")) +
  geom_sf_text(data = veg_forest, aes(label = Zone), size =1.5)+
  geom_sf(data = pts_forest, aes(color = min_tcsa), size = 1.1) +
  theme_minimal()+
  scale_fill_brewer(palette="Greys") +
  labs(color="TCSA") +
  guides(fill = "none") +
  theme(axis.text.x = element_text(angle = 45, size = 4.5)) +
  theme(axis.text.y = element_text(size = 4.5)) +
  theme(axis.title.x = element_blank(),
          axis.title.y = element_blank()) +
  theme(legend.title = element_text(size= 6),
        legend.key.height = unit(4.5, 'mm'),
        legend.key.width = unit(2, 'mm'),
        legend.text = element_text(size = 4))

@@

In order to explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones. Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of the vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of the eco-region map. We observed that there are statistically significant differences in the distribution of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`). The Tukey's HSD results show that TCSA values from the South Temperature Zone (ST) are significantly different from the other two vegetation zones (@fig-gis-tcsa-sites panel C). Looking into this further, we see that there are only six artifacts from the ST zone, with only three of these having TCSA values above 175. The other three have TCSA values lower than 100, similar to the other zones (@fig-gis-tcsa-sites panel B). Given the small number of artifacts with extreme TCSA values in the ST zone, we hesitate to conclude that this is an archaeologically significant pattern of different TCSA values. The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values; this was likely due to the small number of points in several zones (@fig-gis-tcsa-sites panel F). Overall, our results indicate no clear patterns in the distribution of TCSA values across vegetation and eco-regional zones.

@@

```{r}
#| eval: false

@@

# Use reduce() to iteratively join the data frames
tcsa_all <- reduce(dfs, left_join, by = join_col) %>%
            dplyr::select(age_ka.x, 
                          min_tcsa, 
                          Assemblage.x, 
                          raw_materials_top.x, 
                          ecoregion, 
                          Zone) %>%             
  mutate(mis_stage = case_when (
    age_ka.x > 26.5 ~ 'Before' , 
    age_ka.x > 19 ~ 'LGM', 
    age_ka.x < 19 ~ 'After'))

@@

plot_grid(
  tcsa_all_stage_plot,
  tcsa_all_raw_plot,
  tcsa_all_ass_plot,
  tcsa_all_veg_plot,
  tcsa_all_geo_plot,
  ncol = 2,
  label_size = 1
)

# BM combine data frames so we can put them all on a single plot panel

library(ggrepel)
bind_rows(tcsa_all_stage %>% rename(var = mis_stage),
          tcsa_all_raw   %>% rename(var = raw_materials_top.x),
          tcsa_all_ass   %>% rename(var = Assemblage.x ),
          tcsa_all_veg   %>% rename(var = ecoregion ),
          tcsa_all_geo   %>% rename(var = Zone )) %>% 
  ggplot() + 
         aes(x = sample_size, 
             y = CV, 
             label = var) +
  geom_point() +
  geom_text_repel() +
  geom_smooth(method = "lm",
              se = FALSE) +
  labs(y = "CV of Min TCSA", 
       x = "Sample size") +
  theme_minimal()

@@

By comparing the results with the TCSA range from other archaeological and ethnographic cases (@tbl-tcsa-ranges), we were able to determine what role stemmed points played in the technological transition during the Korean Late Paleolithic. Our main questions were: What were the best-fit ballistic probabilities for stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point uses?

@@

Our results show that TCSA values vary between assemblages with few discernible temporal patterns in function (@fig-tcsa-radio-carbon panel A). Contrary to our prediction, the earliest stemmed points do not have more variable TSCA values than subsequent ones. The most striking temporal pattern is simply that stemmed points were produced and discarded most frequently before the LGM (@fig-tcsa-radio-carbon panel B). While we predicted the greatest variation in TCSA values during the LGM, our results show the opposite: variability decreased during the LGM (CV = `r cv_lgm`), in comparison to the previous period (CV = `r cv_bef`). This could be explained by the overall decline in stemmed points usage, perhaps due to a retreat into refugia during the LGM, rather than increased mobility into unfamiliar areas.

@@

Considering the importance of stemmed points for the technological transition during the Late Paleolithic in Korea, our research examines the likely uses for stemmed points by asking three research questions: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point use? We applied the tip cross-sectional area (TCSA) metric because it has been used as a ballistically relevant standard to discriminate different likely use classes of projectile points and it requires only a few measurements on a stone artifact (i.e. width and thickness) to compute the metric. We calculated TCSA for a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites. Then we examined the TCSA values with other variables, including raw materials, weight, radiocarbon dates, and site locations. Drawing on evolutionary theory, we premised that the way of using stemmed points likely reflects the socio-environmental circumstances that people encountered and managed. We examined the possible impact of LGM on stemmed points and their distribution across eco-regional zones.

According to the different weapon-delivery systems that can be inferred from TCSA values, the majority of stemmed points from the Korean Paleolithic were probably used as javelin tips and stabbing spear tips. In general, though, we noted a wide range and also differing distributions of TCSA values in each assemblage. Therefore, we conclude that stemmed points served diverse and not highly patterned roles during the Late Paleolithic. Prior to the LGM, people may have encountered unfamiliar situations and use stemmed points as a multifunctional tool to carry out multiple tasks. During the LGM and following we find production and discard of stemmed points declined, but their ballistic properties remained largely unchanged. Considering that many of our predictions about TCSA were not supported by the data, we speculate that the ballistic aspects of stemmed points were not highly sensitive to global climate dynamics and local geographic variation, but instead were optimal for the wide range of conditions encountered by people during the Late Palaeolithic in Korea.
